\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}

\title{Static type checker for Forth}
\author{Jaanus Pöial \\ Tallinn University of Technology}

\begin{document}

\maketitle

\begin{abstract}
This article develops a mathematical view of static analysis for Forth-like
stack languages. The central object is the symbolic stack effect, a typed and
identity-sensitive abstraction of how a word transforms the stack. Three
operations organize the calculus: composition of linear effects, branch merging
as greatest lower bound in a must-analysis order, and loop summarization by
idempotent invariants. Historically, this line of work grew from algebraic
specifications of stack effects to multiple-effect models, typed wildcards,
polymorphism, inheritance, and later formulations in the language of program
analysis. The article argues that the resulting calculus should be understood as
an intersection of semigroup-style algebra, symbolic execution, abstract
interpretation, and static type systems for low-level code. It is especially
interesting because the three basic operators of program analysis appear here in
an unusually compact form: sequencing becomes effect multiplication, control-flow
join becomes greatest lower bound, and loop reasoning becomes the search for
idempotent summaries.
\end{abstract}

\noindent\textbf{Keywords:} Program analysis; static type checking; symbolic
execution; abstract interpretation; stack languages; Forth.

\section{Introduction}

Forth programs are short, dense, and operationally transparent, but the burden
of reasoning is shifted onto the stack. The meaning of a phrase is often
obvious to an expert and opaque to everyone else. Traditional stack comments of
the form \texttt{( before -- after )} help, but they remain informal,
non-compositional in practice, and usually too weak to express either type
refinement or value identity through stack shuffling.

The natural response is to treat stack effects as semantic objects rather than
mere comments. Once we do that, three classical program-analysis questions
appear immediately.

\begin{enumerate}
\item How do we summarize a linear sequence of words?
\item How do we merge information from alternative branches?
\item How do we compute a stable summary for repeated execution?
\end{enumerate}

In mainstream static analysis these questions correspond to transfer-function
composition, join or meet at control-flow merge points, and fixpoint
computation for loops. In a stack-effect calculus they reappear in a more
algebraic form as effect composition, branch merging, and invariant extraction.

The point of this article is not merely that these analogies exist. It is that
in the stack-language setting they can be made concrete with very small
mathematical machinery. A stack effect is a pair of symbolic stacks; sequencing
is partial multiplication; conditional convergence is a greatest lower bound in
a must-order; loop reasoning is approximated by an idempotent effect or,
equivalently, by a local fixpoint surrogate.

\section{Historical Lineage}

The historical background needed here is short. Early work formalized linear
stack effects algebraically~\cite{poial1990}, then generalized them to multiple
effects so that control-sensitive behavior could be represented~\cite{poial1991}.
Later papers add typed wildcards, polymorphism, and inheritance~\cite{poial2002},
and then recast the same ideas in the language of static analysis,
must-analysis, and loop reasoning~\cite{poial2006}. The 2008 framework
paper~\cite{poial2008} matters mainly because it shows that the calculus is not
just notation: it can be executed as an analyzer.

\section{Mathematical Setting}

\subsection{Types and symbolic values}

Let $(T, \leq)$ be a finite preorder of abstract types. The order is read as
``is at least as specific as'' or, operationally, ``can be used where the other
type is expected.'' For example, one may have
\[
\textit{char} \leq +n \leq n \leq X
\]
\[
\textit{a-addr} \leq \textit{c-addr} \leq \textit{addr} \leq u \leq X
\]
\[
\textit{flag} \leq X
\]
where $X$ is the most general stack-cell type.

A symbolic stack item is a pair
\[
(t, i) \in T \times \mathbb{N}
\]
where $t$ is its type and $i$ is its identity index. We write this as $t[i]$.
The index matters: $x[1]$ and $x[2]$ are distinct symbolic values even if they
share the same type.

Let $\Sigma = T \times \mathbb{N}$. A symbolic stack is a finite sequence in
$\Sigma^*$, with the top of stack on the right. Thus $a\ b\ c$ means that $c$
is topmost.

\subsection{Effects}

A stack effect is a pair of symbolic stacks
\[
e = (L, R) \in \Sigma^* \times \Sigma^*
\]
usually written in Forth notation as
\[
( L \; -- \; R )
\]
where $L$ is the required suffix of the input stack and $R$ is the produced
suffix of the output stack.

The effect should be read relationally. If the current abstract stack has the
form $S L$, then after the effect it becomes $S R$, provided that the symbols
on the boundary can be unified consistently.

We distinguish two special elements:
\[
1 = (\epsilon, \epsilon) \qquad \text{identity effect}
\]
\[
0 = \text{clash / impossible effect}
\]

The identity effect corresponds to ``no change.'' The clash effect is the
failure element produced by incompatible types or impossible unifications.

\subsection{Type meet on comparable boundary symbols}

If two types $t$ and $u$ are comparable, their boundary unification keeps the
more specific one:
\[
\operatorname{meet}_T(t, u) =
\begin{cases}
t & \text{if } t \leq u, \\
u & \text{if } u \leq t, \\
\text{undefined} & \text{otherwise.}
\end{cases}
\]

For symbolic values $t[i]$ and $u[j]$, unification produces a fresh symbol
\[
\mu(t[i], u[j]) = \operatorname{meet}_T(t, u)[k]
\]
with a new identity index $k$. Intuitively, the two local boundary variables
become one shared symbolic value.

This small operation is where static type checking and symbolic execution meet:
the type becomes more precise, while identity is preserved through
substitution.

\section{The Three Core Operations}

\subsection{Operation I: composition of linear effects}

Sequential execution is represented by partial multiplication. At the semantic
level, if $\llbracket e \rrbracket$ denotes the partial transformer induced by
effect $e$, then
\[
\llbracket e_1 * e_2 \rrbracket = \llbracket e_2 \rrbracket \circ \llbracket e_1 \rrbracket
\]
whenever the output boundary of $e_1$ is compatible with the input boundary of
$e_2$.

Operationally, multiplication is computed from the top of stack inward. Suppose
\[
e_1 = (L_1, R_1\, a)
\]
\[
e_2 = (L_2\, b, R_2)
\]
where $a$ is the top output symbol of $e_1$ and $b$ is the top input symbol of
$e_2$. If $\mu(a, b)$ is undefined, then $e_1 * e_2 = 0$. Otherwise let
$z = \mu(a, b)$. Substitute $z$ for both $a$ and $b$, remove the matched
boundary pair, and continue recursively. If one side runs out of boundary
symbols first, the unmatched prefix is carried through.

This gives composition a semigroup flavor: it is associative when defined, it
has identity $1$, and it admits a zero-like clash element $0$. In the simplest
untyped case the structure is closely related to the algebraic objects used in
the early papers~\cite{poial1990}. In the typed symbolic case the multiplication
becomes partial and information-sensitive, but the semigroup intuition remains
exactly the right one.

The key analysis insight is that composition is not mere concatenation of stack
comments. Boundary unification propagates substitutions across the effect being
built. Therefore later uses of a value can refine earlier symbolic positions.
This is why stack shuffling and typed memory access can be handled by one and
the same operation.

\subsection{Operation II: branch merging as greatest lower bound}

For control-flow joins we need an information order. Let $\gamma(e)$ be the set
of concrete stack transformations represented by abstract effect $e$. Define
\[
e_1 \leq e_2 \qquad \text{iff} \qquad \gamma(e_1) \subseteq \gamma(e_2)
\]

Thus $e_1 \leq e_2$ means that $e_1$ is at least as informative as $e_2$; it
describes a smaller set of possible behaviors and therefore gives a stronger
guarantee.

In this order, branch convergence in a must-analysis is modeled by greatest
lower bound:
\[
e_1 \sqcap e_2 = \operatorname{glb}(e_1, e_2)
\]

The direction matters. In may-analysis, one usually merges by least upper
bound, because one wants all behaviors that may happen. Here the goal is the
opposite: we want the strongest effect guaranteed on all paths. Therefore the
correct operator is meet, not join.

For a conditional with branch effects $p$ and $q$, the abstract effect is
\[
(\textit{flag} \; -- \; ) * (p \sqcap q)
\]
up to the exact guard convention of the language. If $p$ and $q$ are
incomparable, the merge is undefined and the analysis correctly rejects the
conditional as lacking a common guaranteed stack behavior.

This is precisely analogous to meet-over-paths formulations in monotone
frameworks and abstract interpretation, except that the property being merged is
not a map from variables to facts but a symbolic summary of stack shape, types,
and identity.

\subsection{Operation III: loop invariants as idempotent effects}

Loops require a stable summary for arbitrary repetition. Let $b$ be the effect
of one loop body execution. An exact must-summary for unbounded repetition is a
fixpoint-like object. Algebraically, the desired summary should not change when
composed with itself:
\[
i * i = i
\]

Such an $i$ is idempotent. As in classical invariant reasoning, idempotence
expresses stability: once the loop invariant has been reached, one more
iteration preserves it.

An exact closure would resemble an infinite meet over repeated compositions,
for example
\[
b^\omega = \bigsqcap_{n \geq 1} b^n
\]
when this object exists in the analysis order. In practice one often uses a
local approximation
\[
\pi^*(b) = b \sqcap (b * b)
\]
which asks what remains true after one iteration and after two iterations, then
keeps only the common guarantee. This is not a full Kleene star, but it plays a
related role: it extracts the repeated, stable part of an effect.

The connection to loop invariants is direct. In Hoare-style reasoning an
invariant $I$ must hold before and after the loop body. In this calculus an
idempotent effect is the stack-transformational analogue of that requirement.
One can therefore read the loop operator in two equivalent ways:

\begin{itemize}
\item algebraically, as the search for an idempotent summary;
\item analytically, as the search for a stable abstract transformer.
\end{itemize}

The 2006 formulation makes an additional subtle point~\cite{poial2006}: all
effects support greatest lower bound and thus form a semilattice, but the
idempotent fragment is especially important for loop reasoning because it is the
stable part of the domain.

\section{Derived Forms for Control Structures}

Once the three operators are available, the usual control forms become small
equations.

For straight-line code:
\[
\llbracket w_1 w_2 \ldots w_n \rrbracket
= \llbracket w_1 \rrbracket * \llbracket w_2 \rrbracket * \cdots * \llbracket w_n \rrbracket
\]

For a two-way conditional:
\[
\llbracket \textit{IF}\ P\ \textit{ELSE}\ Q\ \textit{THEN} \rrbracket
= \llbracket \textit{IF} \rrbracket * (\llbracket P \rrbracket \sqcap \llbracket Q \rrbracket)
\]

For a conditional without explicit else branch:
\[
\llbracket \textit{IF}\ P\ \textit{THEN} \rrbracket
= \llbracket \textit{IF} \rrbracket * (\llbracket P \rrbracket \sqcap 1)
\]
since the false branch has effect $1$.

For a loop body $B$, the idealized repeated summary is an invariant
$\operatorname{Inv}(B)$ satisfying
\[
\operatorname{Inv}(B) * \operatorname{Inv}(B) = \operatorname{Inv}(B)
\]
and intended to approximate arbitrary repetition. Operationally one may use
$\pi^*(\llbracket B \rrbracket)$ or a related fixpoint iteration.

These equations are small enough to fit on a page, but they already capture the
essential analysis pattern for sequencing, branching, and iteration.

\section{Program-Analysis Analogues}

\subsection{Transfer-function composition}

The nearest classical analogue of stack-effect composition is transfer-function
composition in dataflow analysis. Kildall's framework~\cite{kildall1973}
already treats a program point as carrying information transformed by monotone
functions. The difference here is that the abstract state is not an environment
over named variables but a symbolic stack with explicit value identity.

In other words, the calculus is a dataflow framework in which the live carrier
of information is positional rather than lexical.

\subsection{Meet-over-paths and abstract interpretation}

Branch merging by greatest lower bound places the calculus very close to
abstract interpretation~\cite{cousot1977}. The semantic picture is familiar:

\begin{itemize}
\item there is a concrete semantics,
\item an abstract domain of effects,
\item a soundness relation via concretization,
\item monotone operators on abstract information,
\item and stable objects for loops.
\end{itemize}

What is unusual is the direction of approximation. Because the calculus is
formulated as must-analysis, the informative operator at joins is meet. This is
the dual of the more common may-analysis setting. The effect calculus is
therefore best seen as a must-style abstract interpretation over symbolic
stacks.

\subsection{Symbolic execution}

The use of typed wildcards makes the method close to symbolic execution in the
sense of King~\cite{king1976}. A stack item is represented by a symbol rather
than a concrete value; operations transform symbolic expressions; path
combinations retain only what can be guaranteed. The important difference is
that this calculus is intentionally shallow: it does not attempt full path
predicates or arbitrary theorem proving. It concentrates on stack shape, type
compatibility, and identity flow.

That restriction is not a weakness. It is exactly why the method remains small,
compositional, and usable for low-level stack code.

\subsection{Hoare-Floyd invariants}

The loop operator has a clear analogue in Floyd's inductive assertions~\cite{floyd1967}
and Hoare's axiomatic reasoning~\cite{hoare1969}. In both cases the goal is to
find a property preserved by the loop body. Here the property is not a
first-order formula over store variables but a symbolic effect on the stack.
Idempotence plays the role normally played by inductiveness.

This suggests a useful slogan:
\[
\text{Hoare invariant : state predicate :: stack invariant : idempotent effect}
\]
The slogan is imperfect but illuminating.

\subsection{Stack-based bytecode verification and low-level typing}

The calculus also has natural relatives in static typing for stack-based
low-level languages. Bytecode verification for the JVM tracks stack height and
types across control-flow joins~\cite{leroy2003}. Compositional type systems
for stack-based low-level languages~\cite{saabas2006} pursue a similar goal
with more conventional type-theoretic machinery. Type inference for stack-based
languages~\cite{stoddart1993} is an earlier close neighbour.

The distinctive feature of the present calculus, compared with much of that
literature, is the combination of three elements in one object:

\begin{itemize}
\item stack shape,
\item subtype-aware typing,
\item identity-sensitive symbolic flow.
\end{itemize}

That combination is what makes stack shuffling analyzable without abandoning the
compact effect notation that Forth programmers already understand.

\section{Discussion}

The calculus is interesting precisely because it sits between several mature
areas without collapsing into any one of them.

\begin{itemize}
\item It is algebraic because straight-line code is organized by a multiplication-like
law with identity, failure, and idempotence.
\item It is type-theoretic because effects carry subtype information and reject
invalid compositions.
\item It is symbolic because stack items retain identity rather than becoming
anonymous type placeholders.
\item It is an abstract interpretation because branches and loops are handled by
order-theoretic operators.
\end{itemize}

From the point of view of semantics, the core insight is that low-level stack
programs admit a very compact abstract domain. One does not need a large memory
model to recover useful guarantees. For a substantial class of questions,
symbolic stack effects are already the right abstraction.

From the point of view of program analysis, the main theoretical novelty is the
must-style use of greatest lower bounds together with idempotent loop
summaries. In a conventional control-flow graph this would look unusual. In a
stack-effect setting it feels almost inevitable.

There are also clear limitations. A single effect may be too coarse for
path-sensitive reasoning in rich control flow. Exact iteration may require
larger closure operators than $\pi^*$. Data stack and return stack
interactions, heap structure, aliasing, and user-defined control operators may
all force the domain to grow. But these are standard tensions in static
analysis: precision, cost, and compositionality are always in competition.

\section{Conclusion}

Static analysis for stack languages can be presented in a remarkably small
algebra. The entire method rests on three operations:

\begin{enumerate}
\item composition for sequencing,
\item greatest lower bound for branch convergence,
\item and idempotent closure for loop invariants.
\end{enumerate}

Once types and symbolic identities are added, this algebra becomes expressive
enough to function simultaneously as a type checker, a symbolic executor, and a
must-analysis. The broader lesson is that stack languages do not need to remain
outside the main traditions of static analysis. Their structure is different,
but the central ideas, composition, order, and fixpoint stability, are the
same.

\begin{thebibliography}{99}

\bibitem{poial1990}
Jaanus Poial.
\newblock ``Algebraic Specifications of Stack-Effects for Forth Programs.''
\newblock In \emph{1990 FORML Conference Proceedings, euroFORML'90 Conference},
Ampfield, Hampshire, UK, 12--14 October 1990. FIG, 1991, pp. 282--290.

\bibitem{poial1991}
Jaanus Poial.
\newblock ``Multiple Stack-effects of Forth Programs.''
\newblock In \emph{1991 FORML Conference Proceedings, euroFORML'91 Conference},
Marianske Lazne, Czechoslovakia, 11--13 October 1991. FIG, 1992, pp. 400--406.

\bibitem{poial2002}
Jaanus Poial.
\newblock ``Stack Effect Calculus with Typed Wildcards, Polymorphism and Inheritance.''
\newblock In \emph{EuroForth 2002 Proceedings}, Vienna, Austria,
6--8 September 2002, p. 38.

\bibitem{poial2006}
Jaanus Poial.
\newblock ``Typing Tools for Typeless Stack Languages.''
\newblock In \emph{Proceedings of EuroForth 2006}, Cambridge, UK,
15--17 September 2006, pp. 40--46.

\bibitem{poial2008}
Jaanus Poial.
\newblock ``Java Framework for Static Analysis of Forth Programs.''
\newblock In \emph{Proceedings of EuroForth 2008}, Vienna, Austria,
26--28 September 2008, pp. 20--23.

\bibitem{kildall1973}
Gary A. Kildall.
\newblock ``A Unified Approach to Global Program Optimization.''
\newblock In \emph{Proceedings of the ACM Symposium on Principles of Programming Languages},
1973, pp. 194--206.
\newblock doi:\href{https://doi.org/10.1145/512927.512945}{10.1145/512927.512945}.

\bibitem{cousot1977}
Patrick Cousot and Radhia Cousot.
\newblock ``Abstract Interpretation: A Unified Lattice Model for Static Analysis of Programs by Construction or Approximation of Fixpoints.''
\newblock In \emph{Proceedings of the 4th ACM Symposium on Principles of Programming Languages},
1977, pp. 238--252.
\newblock doi:\href{https://doi.org/10.1145/512950.512973}{10.1145/512950.512973}.

\bibitem{king1976}
James C. King.
\newblock ``Symbolic Execution and Program Testing.''
\newblock \emph{Communications of the ACM} 19(7), 1976, pp. 385--394.
\newblock doi:\href{https://doi.org/10.1145/360248.360252}{10.1145/360248.360252}.

\bibitem{floyd1967}
Robert W. Floyd.
\newblock ``Assigning Meanings to Programs.''
\newblock In \emph{Proceedings of Symposia in Applied Mathematics}, Vol. 19,
\emph{Mathematical Aspects of Computer Science}, American Mathematical Society,
1967, pp. 19--32.
\newblock doi:\href{https://doi.org/10.1090/psapm/019/0177187}{10.1090/psapm/019/0177187}.

\bibitem{hoare1969}
C. A. R. Hoare.
\newblock ``An Axiomatic Basis for Computer Programming.''
\newblock \emph{Communications of the ACM} 12(10), 1969, pp. 576--580.
\newblock doi:\href{https://doi.org/10.1145/363235.363259}{10.1145/363235.363259}.

\bibitem{leroy2003}
Xavier Leroy.
\newblock ``Java Bytecode Verification: Algorithms and Formalizations.''
\newblock \emph{Journal of Automated Reasoning} 30, 2003, pp. 235--269.
\newblock doi:\href{https://doi.org/10.1023/A:1024449429949}{10.1023/A:1024449429949}.

\bibitem{saabas2006}
Ando Saabas and Tarmo Uustalu.
\newblock ``Compositional Type Systems for Stack-Based Low-Level Languages.''
\newblock In \emph{Proceedings of the 12th Computing: The Australasian Theory Symposium (CATS 2006)},
Hobart, Australia, 2006, pp. 27--39.

\bibitem{stoddart1993}
Bill Stoddart and Peter J. Knaggs.
\newblock ``Type Inference in Stack Based Languages.''
\newblock \emph{Formal Aspects of Computing} 5, 1993, pp. 289--298.
\newblock doi:\href{https://doi.org/10.1007/BF01212404}{10.1007/BF01212404}.

\end{thebibliography}

\end{document}
